---
title: "Spotted Lanternfly Migration - 2 April 2025"
author: "Clinton Sears"
format: html
editor: visual
---

## Background

The spotted lanternfly (Lycorma delicatula) has been rapidly spreading throughout the northeastern United States since its introduction to Pennsylvania in 2014. While not truly migratory, these invasive insects disperse efficiently by hitchhiking on vehicles, cargo, and outdoor equipment, allowing them to establish new populations across state lines. They pose a significant economic risk because they feed on over 70 plant species, including important agricultural crops like grapes, hops, and fruit trees, causing severe damage by extracting sap and excreting honeydew that promotes sooty mold growth. Their feeding weakens plants, reduces yields, and can lead to plant death, threatening agricultural industries and forest ecosystems throughout the region.

Data were download from the Global Biodiversity Information Facility and include only research grade observations with a CC-BY-NC 4.0 License from iNaturalist.

Citation: GBIF.org (30 March 2025) GBIF Occurrence Download <https://doi.org/10.15468/dl.6ns9jc>, and also cite NJ DOA for quote about spotted lanternfly: <https://www.nj.gov/agriculture/divisions/pi/prog/pests-diseases/spotted-lanternfly/>

## Load Data and Packages, Basic Cleaning

Note that data are already clean. Four observations that were missing lat/lon and state were deleted, leaving 26,364 observations.

```{r}
#| echo: false

slm <- read.csv("C:/Users/clint/OneDrive/Code/Spotted Lanternfly/spotted_lanternfly_clean.csv")

library(tidyverse)
library(sf)
library(maps)
library(patchwork) # For making final, combined viz
library(stringr)  # For str_wrap function
library(grid) # For social media content
library(png) # For social media icons (since I don't have the font)

summary(slm)
# Everything looks fine

table(slm$state)
# It may be helpful to have state code rather than full name...
slm <- slm |>
  mutate(
    state_code = case_when(
      state == "Connecticut" ~ "CT",
      state == "Delaware" ~ "DE",
      state == "District of Columbia" ~ "DC",
      state == "Maryland" ~ "MD",
      state == "Massachusetts" ~ "MA",
      state == "New Hampshire" ~ "NH",
      state == "New Jersey" ~ "NJ",
      state == "New York" ~ "NY",
      state == "Pennsylvania" ~ "PA",
      state == "Rhode Island" ~ "RI",
      state == "Vermont" ~ "VT",
      state == "Virginia" ~ "VA",
      state == "West Virginia" ~ "WV",
    )
  )
table(slm$state_code)
length(slm$state_code)
# No problems with state_code

# Review year data
table(slm$year)
# Let's drop 2025 because it's only March. That will give us 10 years of data
slm <- slm |> 
  filter(year != 2025)
slm$year <- factor(slm$year)

# Create the year_pairs variable
slm <- slm |>
  mutate(
    year_pairs = case_when(
      year %in% c("2015", "2016") ~ "2015-2016",
      year %in% c("2017", "2018") ~ "2017-2018",
      year %in% c("2019", "2020") ~ "2019-2020",
      year %in% c("2021", "2022") ~ "2021-2022",
      year %in% c("2023", "2024") ~ "2023-2024",
      TRUE ~ NA_character_ # Handle any other years
    )
  )

# Review the new variable
table(slm$year_pairs)
```

## Line Chart 2

Updated version of my first chart that is more polished, includes quote from NJ DOA, and text formatted to match the other plots produced below.

::: {.callout-note}
In geom_line, I first used size but that's been deprecated so I updated to linewidth.
:::

```{r}
#| echo: false

year_counts <- aggregate(species ~ year, data = slm, FUN = length)
colnames(year_counts)[2] <- "count"  # Rename the count column

# Create the year plot
year_plot_2 <- ggplot(year_counts, aes(x = year, y = count, group = 1)) +
  geom_line(color = "darkred", linewidth = 1.5) +
  geom_point(color = "darkred", size = 2) +
  geom_text(aes(label = count), vjust = -1, color = "black", size = 3.5) +
  
  # Add text annotation (text box)
  annotate(
    "label", 
    x = 1, # Adjust this x position as needed
    y = max(year_counts$count)/1.5, # Position in the middle vertically
    label = str_wrap("Spotted lanternflies are a significant problem in the US because they feed on sap from over 70 different plant species, including important agricultural crops and trees, weakening them and making them susceptible to disease and death. - New Jersey Department of Agriculture", width = 45),
    hjust = 0, 
    vjust = 0.5,
    size = 3.5,
    lineheight = 0.9,
    fill = "white", 
    alpha = 0.8,
    color = "darkred",
    label.size = NA,
    fontface = "italic"
  ) +
  
  
  labs(title = "iNaturalist Users Track the Lanternfly Uptick",
       subtitle = "Observations by naturalists, biologists, and citizen scientists surged tenfold from 2019 to 2023",
       x = NULL) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 14, hjust = 0),
    plot.subtitle = element_text(face = "italic"),
    plot.margin = unit(c(10,5.5,5.5,5.5), "pt")
  ) +
  coord_cartesian(clip = "off")

# Save the plot with the specified dimensions
ggsave("line_chart_2.png", plot = year_plot_2, width = 6.5, height = 3, units = "in")

# You can also specify other file formats like PDF:
# ggsave("year_plot.pdf", plot = year_plot, width = 6, height = 3, units = "in")

# If you just want to see it in the R graphics window with those dimensions
# (note: this might not always be precise and depends on your window size)
options(repr.plot.width = 6.5, repr.plot.height = 3)
print(year_plot_2)
options(repr.plot.width = NULL, repr.plot.height = NULL) # Reset options
```

## Map Facet 3 - Only 5 Panels

My first map facet was choropleth and just didn't work visually. The 2nd was comprised of 10 panels, one for each year–but that was too much. Five maps was sufficient to tell the story.

```{r}
#| echo: false

# 1. Define the list of states you want to include
selected_states_abbr <- c("CT", "DC", "DE", "MA", "MD", "ME", "NH", "NJ", "NY", "PA", "RI", "VA", "VT", "WV")

# 2. Get spatial data for US states
# Using the 'maps' package directly and then converting to sf
us_states_map <- maps::map("state", plot = FALSE, fill = TRUE)
us_states <- sf::st_as_sf(us_states_map)

# Rename the 'ID' column to 'region' to match the previous logic
us_states <- us_states %>%
  rename(region = ID)

# Convert state names in the spatial data to uppercase for easier comparison
us_states$region_upper <- toupper(us_states$region)

# Create a mapping between state abbreviations and the names in the spatial data
state_abbr_to_name <- data.frame(
  state_code = selected_states_abbr,
  region_upper = c("CONNECTICUT", "DISTRICT OF COLUMBIA", "DELAWARE", "MASSACHUSETTS", "MARYLAND",
                   "MAINE", "NEW HAMPSHIRE", "NEW JERSEY", "NEW YORK", "PENNSYLVANIA", "RHODE ISLAND",
                   "VIRGINIA", "VERMONT", "WEST VIRGINIA")
)

# 3. Filter the spatial data to include only the selected states
eastern_states_spatial <- us_states %>%
  inner_join(state_abbr_to_name, by = "region_upper")

# 4. Filter your slm data for the selected states
slm_eastern <- slm %>%
  filter(state_code %in% selected_states_abbr)

# 5. Create the point map series using ggplot2
point_map_series_3 <- ggplot() +
  geom_sf(data = eastern_states_spatial, color = "gray20", fill = "gray95", linewidth = 0.3) +
  geom_point(data = slm_eastern, aes(x = lon, y = lat), color = "darkred", size = .5, alpha = 0.5) +
  coord_sf(xlim = c(min(slm_eastern$lon) - 1, max(slm_eastern$lon) + 1),
           ylim = c(min(slm_eastern$lat) - 1, max(slm_eastern$lat) + 1)) +
  facet_wrap(vars(year_pairs), ncol = 5, switch = "x") +
  labs(title = "A Decade of Lanternfly Territory Gain",
       subtitle = "Mid-Atlantic hardest hit; Maine the sole New England holdout. Each circle is one observation.") +
  theme_minimal() +
  theme(
    strip.background = element_blank(),
    plot.title = element_text(face = "bold", size = 14, hjust = 0), 
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    panel.grid.major = element_blank(),
    plot.subtitle = element_text(face = "italic")
  )

# Print the point map series
print(point_map_series_3)

# Save the plot with specified dimensions
ggsave("map_facet_3.png", plot = point_map_series_3, width = 6.5, height = 5, units = "in")
```

## Line Facet of Obs by Month

This took a while to get right. The previous two plots use red to show danger/a problem, while this switches to blue and invites the read to take action by using iNaturalist to made observations of spotted lanternflies in their area.

```{r}
#| echo: false

table(slm$species,slm$month,slm$state)
table(slm$species,slm$state)

slm <- slm |>
  mutate(
    state_code_PLUS = case_when(
      state == "Connecticut" ~ "CT [456]",
      state == "Delaware" ~ "DE [506]",
      state == "District of Columbia" ~ "DC [X]",
      state == "Maryland" ~ "MD [3,098]",
      state == "Massachusetts" ~ "MA [46]",
      state == "New Hampshire" ~ "NH [1]",
      state == "New Jersey" ~ "NJ [5,556]",
      state == "New York" ~ "NY [8,629]",
      state == "Pennsylvania" ~ "PA [5,882]",
      state == "Rhode Island" ~ "RI [89]",
      state == "Vermont" ~ "VT [3]",
      state == "Virginia" ~ "VA [1,704]",
      state == "West Virginia" ~ "WV [216]",
    )
  )

# 1. Aggregate the data to count species observations per state and month, excluding DC
state_month_counts <- slm %>%
  filter(state_code_PLUS != "DC [X]") %>% # Exclude DC
  group_by(state_code_PLUS, month) %>% # Group by state and month
  summarise(observation_count = n(), .groups = 'drop')

# 2. Calculate total observations per state
state_totals <- state_month_counts %>%
  group_by(state_code_PLUS) %>%
  summarise(total_observations = sum(observation_count), .groups = 'drop') %>%
  arrange(desc(total_observations))

# 3. Create an ordered factor of state_code based on total observations
ordered_states <- state_totals$state_code_PLUS

# 4. Convert state_code in state_month_counts to an ordered factor
state_month_counts <- state_month_counts %>%
  mutate(state_code_PLUS = factor(state_code_PLUS, levels = ordered_states))

# Create unique single-letter month abbreviations
month_abbr_vector <- c("J", "F", "M", "A", "y", "u", "l", "g", "S", "O", "N", "D") # Using 'y' for May and 'u' for June, 'l' for July, 'g' for August to avoid duplicates

# Recode the month variable as an ordered factor
month_levels <- c("J", "F", "M", "A", "y", "u", "l", "g", "S", "O", "N", "D")
state_month_counts <- state_month_counts %>%
  mutate(month_abbr = factor(month_abbr_vector[month], levels = month_levels))

# 5. Create the line graph faceted by state and showing counts by month
state_month_line_chart <- ggplot(state_month_counts, aes(x = month_abbr, y = observation_count, group = 1)) + # Use month_abbr on x-axis
  geom_line(color = "darkblue", size = 1) + # Use geom_line
  facet_wrap(vars(state_code_PLUS), ncol = 6, switch = "x") + 
  labs(title = "Distribution of Lanternfly Observations from January to December, by State",
       subtitle = "Citizen scientists are most likely to observe adult lanternflies (with distinctive red hindwings) in late summer.\nTotal observations from 2015 to 2024 for each state in brackets.",
       x = "Month",
       y = NULL) + # Added y-axis label
  theme_minimal() +
  theme(
    axis.text.x = element_blank(), 
    strip.background = element_blank(), 
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", size = 14, hjust = 0, color = "darkblue"),
    panel.grid.major.x = element_blank(), 
    panel.grid.minor.x = element_blank(),  
    panel.grid.major.y = element_line(color = "gray95"), 
    panel.grid.minor.y = element_blank(),
    plot.subtitle = element_text(face = "italic")
  ) +
  scale_y_continuous(limits = c(0, 2000)) +
  geom_hline(yintercept = 0, color = "gray60", linewidth = 0.5)

# Print the line chart
print(state_month_line_chart)

# Save the plot with specified dimensions
ggsave("observations_by_state_month_ordered_no_hlines_ylim.png", plot = state_month_line_chart, width = 10, height = 6, units = "in") # Updated filename
```

## Social Media Annotations

Adding this into my caption to make sure folks can connect with me, if needed.

Very helpful blog post: <https://nrennie.rbind.io/blog/adding-social-media-icons-ggplot2/>

Unfortunately, it seems that Font Awesome no longers allows free download of brands logos as a font. I have a bunch of SVGs though...

Code below uses an alternative approach.

```{r}

# Create text grobs for the social media info
library(grid)
library(png)

# Assuming you've converted your SVGs to PNGs
github_icon <- readPNG("C:/Users/clint/OneDrive/Data/social media icons/github.png")
mastodon_icon <- readPNG("C:/Users/clint/OneDrive/Data/social media icons/mastodon.png")

# Create a function to make a social media footer
create_social_footer <- function(github_user = NULL, mastodon_user = NULL) {
  # Create an empty grob
  footer <- grid::gTree(children = grid::gList())
  
  # Add GitHub info if provided
  if (!is.null(github_user)) {
    github_grob <- grid::grobTree(
      grid::rasterGrob(github_icon, x = 0.01, y = 0.5, width = 0.02, just = c("left", "center")),
      grid::textGrob(paste0("github.com/", github_user), x = 0.04, y = 0.5, just = c("left", "center"),
                     gp = grid::gpar(col = "darkblue", fontsize = 8))
    )
    footer <- grid::grobTree(children = grid::gList(footer, github_grob))
  }
  
  # Add Mastodon info if provided
  if (!is.null(mastodon_user)) {
    mastodon_grob <- grid::grobTree(
      grid::rasterGrob(mastodon_icon, x = 0.2, y = 0.5, width = 0.02, just = c("left", "center")),
      grid::textGrob(paste0("@", mastodon_user), x = 0.23, y = 0.5, just = c("left", "center"),
                     gp = grid::gpar(col = "darkblue", fontsize = 8))
    )
    footer <- grid::grobTree(children = grid::gList(footer, mastodon_grob))
  }
  
  return(footer)
}


# Create social media footer
social_footer <- create_social_footer(github_user = "clintonsears", mastodon_user = "clintonsears")
```

## Social Media Footer

Note that gpar(col = "blue") doesn't work. Need to modify original PNG to the color needed, or explore other R packages.

```{r, include=FALSE}

github_icon <- readPNG("C:/Users/clint/OneDrive/Data/social media icons/github_282c79.png")
mastodon_icon <- readPNG("C:/Users/clint/OneDrive/Data/social media icons/mastodon_282c79.png")
# You'll need to add the bluesky icon to your folder
bluesky_icon <- readPNG("C:/Users/clint/OneDrive/Data/social media icons/bluesky_282c79.png")

# Create a function to make a social media footer
create_social_footer <- function(github_user = NULL, mastodon_user = NULL, bluesky_user = NULL) {
  # Create an empty grob
  footer <- grid::gTree(children = grid::gList())
  
  # Current position tracker (x coordinate)
  current_x <- 0.01
  
  # Add Mastodon info if provided (first social media)
  if (!is.null(mastodon_user)) {
    mastodon_grob <- grid::grobTree(
      grid::rasterGrob(mastodon_icon, x = current_x, y = 0.5, width = 0.02, just = c("left", "center"),
                      gp = grid::gpar(col = "blue")),
      grid::textGrob(paste0("@", mastodon_user), x = current_x + 0.03, y = 0.5, just = c("left", "center"),
                     gp = grid::gpar(col = "black", fontsize = 8))
    )
    footer <- grid::grobTree(children = grid::gList(footer, mastodon_grob))
    
    # Update position for next item
    current_x <- current_x + 0.15
  }
  
  # Add GitHub info if provided (second social media)
  if (!is.null(github_user)) {
    github_grob <- grid::grobTree(
      grid::rasterGrob(github_icon, x = current_x, y = 0.5, width = 0.02, just = c("left", "center"),
                      gp = grid::gpar(col = "blue")),
      grid::textGrob(paste0("github.com/", github_user), x = current_x + 0.03, y = 0.5, just = c("left", "center"),
                     gp = grid::gpar(col = "black", fontsize = 8))
    )
    footer <- grid::grobTree(children = grid::gList(footer, github_grob))
    
    # Update position for next item
    current_x <- current_x + 0.20
  }
  
  # Add Bluesky info if provided (third social media)
  if (!is.null(bluesky_user)) {
    bluesky_grob <- grid::grobTree(
      grid::rasterGrob(bluesky_icon, x = current_x, y = 0.5, width = 0.02, just = c("left", "center"),
                      gp = grid::gpar(col = "blue")),
      grid::textGrob(bluesky_user, x = current_x + 0.03, y = 0.5, just = c("left", "center"),
                     gp = grid::gpar(col = "black", fontsize = 8))
    )
    footer <- grid::grobTree(children = grid::gList(footer, bluesky_grob))
  }
  
  return(footer)
}

# Create social media footer with all elements
social_footer <- create_social_footer(
  github_user = "clintonsears", 
  mastodon_user = "clintonsears",
  bluesky_user = "@clintonsears.bsky.social"
)

# To view the footer (uncomment to test)
grid.newpage()
grid.draw(social_footer)
```

## Final Viz

4th version...

Previous "final" versions incorporated plots that simply didn't work out, so this is the actual final visual.

::: {.callout-note}

Review the final PDF or PNG to see properly formatted visual.

:::

```{r, include=FALSE}

# Create the combined plots first
combined_plot <- year_plot_2 /
                    point_map_series_3 / 
                    state_month_line_chart +
                    plot_layout(heights = c(.7, 1, 1),
                                ncol = 1,
                                guides = "collect") &
                    theme(plot.margin = margin(0.5, 0.5, 0.5, 0.5, "cm"))

# Create an annotation grob for your social media footer that will be outside the plot layout system
library(grid)
library(gridExtra)

# Create a gtable for holding the social media content
social_gtable <- arrangeGrob(
  social_footer,
  bottom = textGrob("", gp = gpar(fontsize = 8))
)

# Add title and annotations to create final_plot_3 with proper caption
final_plot_3 <- combined_plot +
  plot_annotation(
    title = "**From Introduction to Infestation:**<br>**The Spotted Lanternfly's Population Explosion**",
    subtitle = "<br>Introduced to Pennsylvania in 2014, the spotted lanternfly (*Lycorma delicatula*) has <br>rapidly spread throughout the northeastern US, posing a significant threat to agriculture<br> and ecosystems. This visualization depicts its spread using research-grade observations<br> submitted to iNaturalist and accessed via Global Biodiversity Information Facility.",
    caption = "Data Sources: GBIF.org (30 March 2025) GBIF Occurrence Download https://doi.org/10.15468/dl.6ns9jc and \nNew Jersey Dept. of Agriculture https://www.nj.gov/agriculture/divisions/pi/prog/pests-diseases/spotted-lanternfly/",
    theme = theme(
      plot.title = ggtext::element_markdown(size = 24, hjust = 0, color = "darkred"), 
      plot.subtitle = ggtext::element_markdown(size = 14),
      plot.margin = margin(1, 1, 1, 1, "cm")
    )
  ) &
  theme(plot.margin = margin(0.7, 0.2, 0.7, 0.2, "cm"))

# Print the combined plot
print(final_plot_3)

# Save the plot to a temporary file
temp_file <- tempfile(fileext = ".png")
ggsave(temp_file, plot = final_plot_3, width = 8.5, height = 11.7, units = "in")

# Now read it back in as a single image
library(png)
plot_img <- readPNG(temp_file)

# Create a new plot with the image and the social footer
png("spotted lanternfly_FINAL.png", width = 8.5, height = 12, units = "in", res = 300)
grid.newpage()
# Draw the main plot image (takes up 98% of height)
grid.raster(plot_img, width = unit(1, "npc"), height = unit(0.98, "npc"), 
            just = c("center", "top"), y = unit(1, "npc"))
# Draw the social media footer at the bottom (takes up 2% of height)
pushViewport(viewport(y = unit(0.01, "npc"), height = unit(0.02, "npc"),
                      x = unit(.94, "npc"),
                     just = c("center", "bottom")))
grid.draw(social_footer)
popViewport()
dev.off()

# PDF process is a bit different...
# After a lot of fuss I could not get PDF to save correctly.
```
